---
title: Liveness Gap Analysis
category: algorithms
slug: liveness-gap-analysis
nav_order: 5
status: current
---

\section{Liveness Gap Analysis}

\subsection*{Context}

This document maps each condition in the Retry Policy Invariant ($RPI$) to
its implementation status. It supersedes the prior gap analysis, which
incorrectly classified $\Pi_2$ as unimplemented.

The gap is \textbf{narrower than previously stated}. The primary open gap
is $\Pi_4$ at the supervisor boundary.

\subsection*{What Is Currently Established}

\begin{itemize}
  \item \textbf{Deadlock Freedom} ✅ — scratchpad semantics, no locks held
  \item \textbf{Bounded Termination} ✅ — \texttt{max\_turns=200}, \texttt{\_WORKER\_RUN\_TIMEOUT=1200}
  \item \textbf{Probabilistic Convergence} ✅ — under independence assumption
  \item \textbf{$\Pi_2$ Worker Liveness} ✅ — \texttt{\_wait\_for\_run\_completion} with \texttt{asyncio.TimeoutError}
  \item \textbf{$WF_{\text{vars}}(WorkerCompletes)$} ✅ — sound, backed by timeout mechanism
\end{itemize}

\subsection*{$\Pi_1$ — Supervisor Persistence (Soft)}

The \texttt{while turn\_count < max\_turns} loop in \texttt{process\_conversation}
provides a structural retry container. If the LLM emits a tool call, the loop
re-enters delegation. This is reinforced by \texttt{L4\_EXECUTION\_LOOP} instructions
mandating re-delegation on unverified scratchpad entries.

However, $\Pi_1$ is \textbf{instruction-dependent, not algorithmically hard}.
A model that decides to synthesize prematurely — outputting a terminal text reply
rather than a tool call — breaks persistence silently. The loop exits cleanly
because \texttt{get\_function\_call\_state()} returns empty, which is an
indistinguishable success signal.

\textbf{Status:} ⚠️ Soft — structurally present, behaviourally contingent.

\subsection*{$\Pi_2$ — Worker Liveness (Implemented)}

\texttt{\_WORKER\_RUN\_TIMEOUT = 1200} in \texttt{DelegationMixin} is the
concrete mechanism. \texttt{\_wait\_for\_run\_completion} raises
\texttt{asyncio.TimeoutError} on expiry. This is caught in
\texttt{handle\_delegate\_research\_task}:

\begin{verbatim}
except asyncio.TimeoutError:
    execution_had_error = True
    final_run_status = "timed_out"
\end{verbatim}

The error is submitted as tool output and the supervisor loop continues.
Workers cannot run indefinitely. The fairness annotation $WF_{\text{vars}}(WorkerCompletes)$
in the TLA+ spec is sound against this mechanism.

\textbf{Status:} ✅ Implemented.

\subsection*{$\Pi_3$ — Fair Scheduling (Infrastructure-Dependent)}

Worker runs are created via \texttt{create\_ephemeral\_run} and polled at
\texttt{\_WORKER\_POLL\_INTERVAL = 2.0s}. Whether pending runs are selected
fairly depends on the task queue infrastructure (Celery or equivalent)
that is not visible in this codebase.

\textbf{Status:} ⚠️ Infrastructure-dependent — unverified at this layer.

\subsection*{$\Pi_4$ — Recovery at Supervisor Boundary (Missing)}

This is the primary open gap. When \texttt{turn\_count} reaches \texttt{max\_turns=200}
in \texttt{process\_conversation}, the current behaviour is:

\begin{verbatim}
yield json.dumps({
    "type": "error",
    "content": "Maximum conversation turns exceeded"
})
\end{verbatim}

followed by \texttt{failed\_at} stamp and exit. There is no:

\begin{itemize}
  \item Re-planning with reduced scope
  \item Escalation to a human or higher-level supervisor
  \item Partial synthesis with explicitly declared uncertainty
  \item Any signal to the caller that recovery is possible
\end{itemize}

Note: at the \textbf{worker level}, partial $\Pi_4$ exists — a timed-out worker
submits an error as tool output and the supervisor may choose to re-delegate.
But the supervisor itself has no equivalent when it hits its own bound.

\textbf{Status:} ❌ Not implemented at supervisor boundary.

\subsection*{Correlated Failure Risk}

\texttt{get\_delegated\_model()} returns the same model for all workers.
If the model fails consistently on a task type, $p$ is not stationary across
delegations. The $P(\text{success by } n) = 1 - (1-p)^n$ bound is optimistic
under correlated failure. No mitigation currently exists.

\textbf{Status:} ❌ Unmitigated.

\subsection*{Revised Gap Summary}

\begin{itemize}
  \item $\Pi_1$: structurally present, behaviourally soft ⚠️
  \item $\Pi_2$: implemented ✅
  \item $\Pi_3$: infrastructure-dependent ⚠️
  \item $\Pi_4$: missing at supervisor boundary ❌
  \item Correlated failure: unmitigated ❌
\end{itemize}

The system is closer to live than the prior analysis stated.
The one hard gap is $\Pi_4$: the supervisor can exhaust its turn budget,
produce nothing, and stop cleanly without any recovery signal.