---
title: Liveness Theorem
category: algorithms
slug: liveness-theorem
nav_order: 7
status: proposed
---

\section{Liveness Theorem}

\subsection*{Status}

\textbf{Proposed — closer to verified than previously stated.}

$\Pi_2$ is fully implemented. $WF_{\text{vars}}(WorkerCompletes)$ is sound.
The theorem is blocked on $\Pi_4$ (supervisor recovery handler) and
the hardening of $\Pi_1$ (premature synthesis guard).

\subsection*{Dependencies and Current Status}

\begin{itemize}
  \item \textbf{DF} — Deadlock Freedom ✅ verified
  \item \textbf{TB} — Bounded Termination ✅ verified ($D_S=200$, $T_W=1200\text{s}$)
  \item \textbf{PC} — Probabilistic Convergence ✅ verified (under independence assumption)
  \item \textbf{$\Pi_1$} — Supervisor Persistence ⚠️ soft
  \item \textbf{$\Pi_2$} — Worker Liveness ✅ implemented
  \item \textbf{$\Pi_3$} — Fair Scheduling ⚠️ infrastructure-dependent
  \item \textbf{$\Pi_4$} — Supervisor Recovery ❌ not implemented
\end{itemize}

\subsection*{Theorem}

Under $DF \wedge TB \wedge PC \wedge RPI$:

\[ \Diamond\ \text{Success} \]

If work is submitted, the supervisor eventually produces a successful result
or explicitly invokes recovery — not a silent failure.

\subsection*{Proof}

\textbf{Step 1 — No starvation in $Awaiting$} ✅ \textit{holds today}

By $\Pi_2$ (\texttt{\_WORKER\_RUN\_TIMEOUT = 1200s}), every spawned worker reaches
a terminal state within bounded time. The supervisor cannot idle indefinitely
in $Awaiting$:

\[ \square \left( Awaiting \rightarrow \Diamond_{\leq 1200s}\ Evaluating \right) \]

\textbf{Step 2 — Persistence through evaluation} ⚠️ \textit{soft today}

By $\Pi_1$, whenever $f(C) = incomplete$ and $d < D_S$, the supervisor
re-enters $Delegating$. This holds when the LLM follows \texttt{L4\_EXECUTION\_LOOP}
instructions. It breaks silently if the LLM synthesizes prematurely.

Hardening $\Pi_1$ with a tool-call guard would make this step unconditional.

\textbf{Step 3 — Convergence under retry} ✅ \textit{holds under independence assumption}

By $PC$, each delegation succeeds with probability $p > 0$.
After at most $D_S = 200$ delegations:

\[ P(\text{success by } D_S) = 1 - (1-p)^{200} \]

This bound is optimistic if failures are correlated across a single model.

\textbf{Step 4 — Recovery at depth bound} ❌ \textit{not implemented}

If all $D_S$ delegations fail, the current system yields a hard error.
By $\Pi_4$ (proposed), the system would instead invoke a recovery policy —
partial synthesis, scope reduction, or escalation — preventing silent failure.

Until $\Pi_4$ is implemented, $\Diamond\ \text{Success}$ does not hold
at the supervisor boundary when all turns are exhausted.

\textbf{Step 5 — No deadlock} ✅ \textit{holds today}

By $DF$, $E = \emptyset$ at all reachable states.

\subsection*{Current Liveness Guarantee}

The system today guarantees:

\[ \Diamond\ Terminated \]

where $Terminated$ includes both successful completion and the hard-error
max-turns exit. This is weaker than $\Diamond\ \text{Success}$.

The gap between $\Diamond\ Terminated$ and $\Diamond\ \text{Success}$ is exactly $\Pi_4$.

\subsection*{What Closes the Theorem}

\begin{itemize}
  \item Implement a recovery handler in \texttt{process\_conversation} at
        \texttt{turn\_count = max\_turns} — closes $\Pi_4$
  \item Add a premature-synthesis guard in \texttt{process\_conversation} —
        hardens $\Pi_1$
  \item Verify Celery queue fairness semantics — closes $\Pi_3$
\end{itemize}

\[ \therefore \text{Liveness is } \Diamond\ Terminated \text{ today, and } \Diamond\ \text{Success} \text{ after } \Pi_4 \text{ is implemented.} \]

\qed